Detecting skill level and mental template in biface morphology: Archaeological and experimental insights

### 2022-05-20 manuscript/manuscript.Rmd
added another paragraph in discussion.

The biface experimental replicas used in this study comprised two sub-collection. The first sub-collection includes 10 bifaces knapped by three expert knappers, including Bruce Bradley (n=4), John Lord (n=3), and Dietrich Stout (n=3) [@stout2014]. The second sub-collection is produced from a 90-hour handaxe knapping skill acquisition experiment [@bayani2021; @pargeter2019; @pargeter2020], where 30 adults with no previous experience in knapping were recruited from Emory University and its surrounding communities and requested to make 132 bifaces in total. Among these 30 adult participants, 17 have gone through multiple one-to-one or group training sessions that amounted to 89 hours in maximum, while the remaining 13 were assigned to the controlled group, where no formal training is given.

@@

Training is important in terms of its immediate effects (pre-training vs. post-training) in both skill level and mental template, however, 90 hours of training is still not enough for novice to reach the reach the skill level as reflected in expert knappers or Late Acheulean toolmaker.

@@

Maybe also something about how this study demonstrate the potential of archaeological data reuse and its contempoary relevance [information age+pandemic+ethical research]?

>>>>

The biface experimental replicas used in this study comprised two sub-collection. The first sub-collection includes 10 bifaces knapped by three expert knappers, including Bruce Bradley (n=4), John Lord (n=3), and Dietrich Stout (n=3) [@stout2014]. The second sub-collection is produced from a 90-hour handaxe knapping skill acquisition experiment [@bayani2021; @pargeter2019; @pargeter2020], where 30 adults with no previous experience in knapping were recruited from Emory University and its surrounding communities and requested to make 132 bifaces in total. Among these 30 adult participants, 17 have gone through multiple one-to-one or group training sessions that amounted to 89 hours in maximum, while the remaining 13 were assigned to the controlled group, where no formal training is given. As part of the preparation efforts, the experimental team spalled the Norfolk flints acquired through Neolithics.com into flat blanks of similar size and shape for training and assessments.

@@

In terms of our second research question, this study shows that training do have an immediate intervention effects (pre-training vs. post-training) in both PC1 (skill level) and PC2 (mental template). Nontheless, once the training has been initiated, its effects across different assessments on both dimensions are rather unconsipicous. This finding corroborates what has been suggested in Pargeter et al. [-@pargeter2019] that 90 hours of training for handaxe making is still not enough for novices to reach the skill level as reflected in expert knappers, considering the massive social support involved in the experiment set up including the direct and deliberate peadagogy and the simplified raw material procurement and preparation procedures. This follow-up project further adds the samples produced by Late Acheulean toolmaker as a new benchmark to deepen our understanding on this issue. It is noteworthy that how constrained is the Boxgrove assemblage morphological variation as measured by both PC1 and PC2 even when compared with the modern expert group (**Figure** \@ref(fig:GeneralPCA1)), especially given the fact that it has the largest sample size among all studied group. Some potential explainations for this phenomenon include 1) the strong idiosyncrasy of individual expert knappers shaped by their own unique learning and practice experience; and/or 2) the present day skill shortage of our expert knapper as compared with Boxgrove knappers despite their multiple years of knapping practice [@milks2019].

@@

Another contribution that we would like to highlight here is that this research project demonstrates the potential of reusing old archaeological data in digital format to address novel research questions. In this paper, the main source of archaeological data is a collection of photos produced and curated by one of our co-authors (R. Iovita) more than 10 years ago, and the morphological variation data of the experimental collection are also derived from photographs instead of remeasurements of the original replicas. Given the irreversible nature of archaeological excavation, the digitized data, be it text, pictures, or videos, often become the sole evidence that are available for certain research questions. Yet, it has been widely acknowledged that the reuse of archaeological data has not received enough attention among researchers in our discipline [@faniel2018; @huggett2018; @moody2021]. Among many reasons of preventing archaeologists from reusing published and digitized data [@sobotkova2018], the lack of a standardized practice of and motivation for data sharing is a prominent one [@marwick2018]. As stated in the method section, we addressed this issue through sharing the raw data and the code for generating the derived data on an open-access repository. Another major and legitimate concern of archaeological data resue is their quality. In terms of this aspect, we do acknowledge the limitations of relying on photos when it comes to more detailed technological analysis of stone artifacts, however, our paper shows that finding the approriate research questions given the data available is key to revealing new novel insights into the studied topic. Moreover, we believe that this type of research has a strong contempoarary relevance due to the continued influence of the COVID-19 on fieldwork-related travel and direct access to archaeological artifacts [@balandier2022; @ogundiran2021].
